\subsection*{The guardian of the \textit{legundi} fruits}

During a riot they fled, the mother and the father, and their child they carried on their backs, and he cried; and the father said: ``I just said, don't carry him.''

After that they went inland and put him down on an anthill and left, but later they returned to him again and put him down on the root spur of a tree.

Then they went away, over seven hills and seven rivers, but they still heard him cry, and the mother wanted to return to him but the father did not want that. He said: ``If you return to him, then I will keep going.''

And that child ate only eggplants, and when he had grown up, he once left the Itio river, and he met an old woman, the well guardian of the Western King, and he, the child, said: ``Grandmother, maybe you (are) a witch,'' and she said: ``Child! I am not a witch.''

\newpage
And she asked: ``You here, where are you coming from?'' And he said: ``I have come from the Itio River seawards,'' and the child asked for coconuts: ``Grandmother! do you maybe have coconuts?'' And she said: ``There are coconuts here, but maybe you don't climb?'' And he said: ``Grandmother, of course I know how to climb!'' Then he climbed the coconut tree, retrieving seven coconuts, four for her, three for him. Then they ate and after that she said: ``Child! shall we go rowing?'' And he said: ``That's good grandmother, let's go rowing.'' Then they went rowing and inland they met the seven maidens of the Western King, fetching water, and they said: ``Hey grandmother, our brother.''

Then they went bathing, and they, the seven maidens, said: ``Grandmother! let's push him into the well,'' and they pushed him in. But he floated and one half of his body was silver, and the other half was gold, his hair was like that as well.

Then the seven children of the Eastern King walked southwards, and after that they served them betel, and their brother came out, but they fainted, and he leapt onto their heads so that they were startled. And the Western King asked: ``Now you go down to play ball, or will you let the cocks fight?'' And they said: ``Now we'll go down, we'll play ball, no, we'll let the cocks fight, no, we'll just go for a walk.''

And then the king fired a cannon, so they returned, and they again climbed up and they did their hair in a wreath and they chewed betel. Afterwards the Western King fired a cannon, so they returned, and after that they went down again to go for a walk, they did their hair in the manner of the Muslims.

And they served them betel, their betel box was a \textit{nahinara} and they served them betel and after that, the Eastern King fired a cannon, therefore they returned.

And the \textit{legundi} guard approached his lover and he handed her a bunch of \textit{legundi} fruits, but she did not accept it anymore, and he said: ``Why don't you take it?''

And the children of the Western King went down and they said: ``Father! a moment ago we went up and the \textit{legundi} guard came and gave his lover a bunch of \textit{legundi} fruits, but she didn't take it.''

And once again they went up, the youngest took her broom with her, and she, going up, went straight to the lower part of the \textit{legundi} fruit tree, and she swept there. And he came once again and he shook the remaining rain water from the \textit{legundi} tree so that it fell on her, and she said: ``Hey you, you're harassing me!''

And she asked: ``Are these your \textit{legundi} fruits?'' And he said: ``These are my \textit{legundi} fruits.'' ``If these are your \textit{legundi} fruits, from where do you come here to greet me?'' And he said: ``I'm coming from Limau ma ituo'a down to greet you,'' and she said: ``If that's the case, then your \textit{legundi} fruits are no longer yours.''

He went away and he told it to his father: ``Father! I just went and they said that my \textit{legundi} fruits are no longer mine! They asked me: `Where did you come from to greet us?' And I said: `From Limau ma ituo'a I came down to greet you.'{''} And the father said: ``Thus your \textit{legundi} fruits are no longer yours.'' And he said: ``If that's the case, then let it be so.''

And he wed the child of the Western King too, and the child of the Eastern King wed the son of the Western King and they celebrated her seven nights and seven days.
