\subsection*{The \textit{bibiti} fish}

Once they went fishing with a net -- a mother and her child -- and they caught a \textit{bibiti} fish.

The mother said: ``Let's strike it dead.'' And she said: ``Don't strike it dead, let it be my plaything,'' then she placed it up in a \textit{gohapanata} tree. But look! she went to see it and the fish had grown/multiplied to fill out the \textit{gohapanata} tree completely, and the child said: ``Father! please make a proa, so that I can put my \textit{bibiti} fish in there.''

Then she put it in there, but look! She looked there and saw that they had filled it.

And she said: ``Father, you still have to weed my garden, because those of my friends are already weeded.''

And he said: ``I'll weed it, but you go landwards first,'' and she went landwards, and look, when she came landwards they had already weeded it and the roots had been pulled out.

And she went seawards and she said: ``Father, don't go inland, I'm already done with the weeding.''

And at night someone came to court her, and she refused, saying: ``I didn't come here to court.'' And he said: ``If that's the case, will you just do my hair?'' but she refused. ``Will you just oil me?'' but she refused. ``But just peel betel nuts for us, so that we can chew betel.'' But she refused. Afterwards, he descended. Her betel box was a betel bag (which is carried on the back).

And then the sun shone on him and she saw him well and she said: ``Come here so that I can make up your hair,'' but he refused, saying: ``Don't do that! Because then your father and your mother will be angry with you,'' and she refused and followed him.

And when he descended from the top of a felled tree, she climbed up on its lower end, but when he climbed up, she descended.

Then they followed seven rivers and crossed over seven mountains, but when he climbed, she descended, and when he descended, she climbed.

And they found a river over which a liana hung, then she awaited him, and they chewed betel, and he said: ``Return! Otherwise, you will not find your mother and your father anymore,'' and she said: ``I don't want to, I will follow you,'' then he said: ``Then close your eyes,'' and she closed them, but look! she opened them at the center of the mountain top. A liana hung there and she slept and she spread out her hair over two river bays. And fish filled her hair up to the top, tight like a case net that is completely full.

Once someone fished with a net, and caught her hair, and he said: ``Here now, whose hair is this?'' He said: ``Come, a little bit more towards the bank,'' and they went inland and she slept on a liana and they loaded her into the proa. And they went and there the wives of the king were angry. They said: ``You collect women; now your wives surely number seven already.'' And they said: ``She is beautiful; well, let's poop!'' Then they pooped, but the smell of her excrement was sweet, and they said again: ``She is beautiful; come, let's fart,'' and they farted, but look! the smell of her fart was sweet.

And they said again: ``Come, let's follow the river down to the sea with a raft,'' and they followed the river with a raft. 
They, the seven, went first, and they crashed against the bank and went down into the water at once. And she took a raft too. And she also drifted away with it, and she drifted inland, returning.

And the king said: ``I will marry you.'' And she said: ``If you want to marry me, I first must stab one of your slaves dead.'' And she stabbed him, so that he fell to the ground and he was bound with taro leaves and they left him alone in another house. But look! after seven nights he played the flute, and he was alive again. And half of his face was silver and the other half was gold, half of his hair was silver, the other half gold, his upper teeth were silver and the lower ones were gold.

And the king said: ``Stab me too,'' and she said: ``No, because then you will also become half silver and half gold.'' But he withstood her, so she stabbed him; and he became half and he did not return anymore.
