\subsection*{The \textit{goda} spirit}

There was someone, a married woman. Her husband was a hunter, he went and called the dogs, and he urged his mother, saying: ``Mother! Don't let your daughter-in-law stay on her own, because then the \textit{goda} spirit will seize her.''

And she said: ``No, I won't leave her alone, just go.'' And he left, and was gone. And she, the woman, said: ``Child, let's bathe and wash our hair with coconut milk.'' And the daughter-in-law said: ``That's good, let's bathe, and wash our hair with coconut milk.'' And they bathed and washed their hair; and after that, they embarked in a proa. And the daughter-in-law thought and said: ``Mother-in-law, let's sleep,'' and she said: ``That's good.'' And they lay down, so the mother-in-law fell asleep. The wife of the hunter went to cut bamboo, she put the bamboo close to the \textit{goda} spirit and the \textit{goda} spirit seized her. In this moment, the husband came and he asked his mother, saying: ``Mother, how is your daughter-in-law?'' And she said: ``I was sleeping and she chopped bamboo and the spirit seized her.''

And he said: ``I just said, don't let her alone,'' and he went away to look for her and above her ankle the spirit had seized her.

And he took her silver and her hairpins and put them on her but look! He looked at her and the spirit had already drowned her.

And he returned to his house and afterwards he said: ``I will follow her,'' and he went, returning to her.

Look, after seven days there at their place they were playing the violin, and playing the flute, and the mother looked for them, and look, they came up to the door, and they had returned again.
