\subsection*{The ants}

There was an old woman and her bananas were eaten by ants. The old woman said: ``These bananas of mine, who has eaten them?'' And she said: ``They have eaten my bananas, so I will continue guarding them,'' and she guarded them.

Look! After that the two ants ate; and she said: ``The ants down here are eating my bananas, I will strike them dead.'' And the ants said: ``Grandmother, don't strike us dead, but take us to your home,'' and she took them to her house, and after that she said: ``Old man, a moment ago my bananas were eaten by ants, and look, I wanted to strike them and they said: ``Grandmother! Don't strike us dead, but take us to your house.'' And he, the old man, said: ``So you'll raise them.''

And once the old woman went to her garden, to weed her garden. And she returned home and the ants asked: ``Old woman! Are you done weeding your garden?''

And she said: ``Not yet, I am not yet done with it,'' and the ants said: ``Tomorrow, we will follow you.'' And the old woman said: ``Go ahead, tomorrow we will go.'' And the next day they went, they went and weeded and at once everything was done. And they returned, they went and she told the old man, saying: ``Old man, a moment ago we went inland and they weeded and at once it was done.'' And the ants said: ``Old people, let us search you for lice,'' and they said: ``Grandchildren! Go ahead, search us for lice,'' and they searched them for lice and they said: ``Look up!'' And they looked up, and they, the ants, threw lime at them, and the eyes of the old man hurt so that he screamed; and the ants ran away, and they went and climbed a tree and the old man followed them.

He saw them climbing a tree and the old man said: ``I will strike you dead! We haven't done anything to you, but you ants threw lime in our eyes.'' And he cut them down, but when the tree fell down, they jumped to another tree; then he hacked it down, so when it fell, they jumped to another tree; then he hacked it down, so when it fell, they jumped to another tree; and he got tired and a flying palace came down and the ants embarked the flying palace and they came to the ground with it, from above they fell with it and they sank down.

So the old people said: ``Just come down, we won't do anything to you.'' In the flying palace they played the bamboo guitar, and they played the violin, and they played the flute and they beat the drum and had a jinn feast.

And the young men wanted to marry them, and they said: ``If you want to marry us, you have to impale bamboo leaves, you must first throw them,'' and they threw them and speared them up; after that they came and they wanted to marry them, and they said: ``You first have to throw bamboo leaves!'' And they threw them and the leaves were speared, then they came and they wanted to marry them, and they said: ``You first have to throw bamboo leaves (again),'' and they threw them and they speared them up, then they came and they wanted to marry them, and they said: ``You first have to throw bamboo leaves,'' and they threw them and they speared them up.

Then they came and they wanted to marry them, and they said: ``You first have to spear bamboo leaves, you have to throw them,'' and they threw them, raising them, then they came and they wanted to marry them, and they said: ``You first have to throw bamboo leaves,'' and they threw and speared them up.

And the son of the Western King came too and said: ``If you want, I will marry you,'' and they said: ``You have to throw bamboo leaves first, then you (can) marry us.'' And he threw them, but he missed, and he married them. Then their drums and their gongs were beaten.
