\subsection*{The older and the younger brother}

There were two people, an older and a younger brother. The older one called for the dogs (was a hunter) and the younger one went beating sago. The older one went to call the dogs and to his wife he said: ``If in a little while your brother-in-law comes here and he asks for sago bread, then you have to give it to him.''

And the wife said: ``Just go, I'll give it to him,'' and he went and when he was gone, his younger brother came. And he came and asked for sago bread, saying: ``Sister-in-law, can I eat some sago bread? I'm hungry.'' And she said: ``Sago bread, where here do I have that?'' and his bowels rumbled.

And he, the brother-in-law, was ashamed and he went away, climbing seven mountains.

And the older brother came back and he asked his wife: ``How is your brother-in-law?'' And she said: ``Do I know about him?'' And he said to himself: ``It is like this, he has asked for sago bread and she has not given it to him, so he left.''

And he followed him, climbing seven mountains and he hit a tree root and stopped. And then they both stopped, they chewed betel and the older one asked: ``Why did you go away?'' And he said: ``It's nothing, I asked your wife for sago bread and she did not give it to me.''

And the older one said: ``Come here, let's return!'' And he said: ``I won't return.''

And they turned their backs to each other, he, the oldest, returned to their house, and he, the younger one, went to foreign lands, and he went and came to an old woman.

And he sent the dog, saying: ``Dog, go landwards and if the grandmother is a witch, then tell me!''

And it went landwards and returned and said: ``She is not a witch,'' and he also sent his ring and it came seawards and it said: ``She is not a witch.''

And he shouted and showed himself and he said: ``Grandmother! Grandmother!'' and she said: ``Hey, I don't have a grandchild.'' And he said: ``Grandmother, maybe you are a witch, but don't take me!'' And she said: ``I'm not a witch but a real human.''

And they went to her house and she cooked rice and they ate, and after they had eaten, it had become night, so they slept.

And at midnight, a worm fell into their house, and he was startled by it and said: ``Certainly that's her'' and he called, saying: ``Grandmother, what is it?'' and she said: ``There is a worm that has fallen down.'' And she lighted a torch, and it was a worm that had fallen.

And the next day seven daughters of the Eastern King came walking southwards; they went southwards and they said: ``Grandmother, your smell is sweet, like the smell of supernatural beings.''

And she said: ``Grandchildren, there is nothing -- in my house jasmine flowers smell like this,'' then they descended into the water.

And the old woman asked the younger brother: ``Which ones of those seven do you like?'' And he said: ``The last one.'' And the king fired a cannon, so they returned.

When they arrived at the top their drums and gongs were beaten; and he asked: ``Grandmother, what feast is this?'' And she said: ''People say, they stab with tops of \textit{banahana} plants at each other.'' And he said: ``Grandmother, if they bleed, will we go watch them?'' And they went and at the end of the path they sat down. And she played and it\footnote{The referent is unclear.} fell down on his head and also on his lap. And he kicked as well, so it fell down on her head and on her lap too.

\largerpage
Then they returned, but look, after that they came again, the seven and they were disguised. And he, the young man, hid under a jasmine bush, and the mask of the youngest hit him by accident and they descended into the water and he hid it.

Look, the older sisters went and her mask was gone then the king fired a cannon and they, the older sisters, returned already.

And the old woman said: ``Don't cry, go up,'' and she went up and said: ``It is my mask that he is showing,'' and she married him.

And the king sent to fetch them, her and her husband, and they descended, fetched them and brought them up and they feasted them for seven nights and seven days.

And the older brother followed, and he also reached the old woman, then they descended again, the six sisters, and they were disguised.

And the old woman asked: ``Grandson, of the six daughters of the king a moment ago, who do you desire?'' And he said: ``That one in the middle,'' then the king fired a cannon, so they returned again. After that they descended again and they were disguised; and he hid under a jasmine bush and the mask of the middle one accidentally hit him and they dived into the water and he hid it.

Look, the king fired a cannon and her mask was gone and the older sisters returned and the middle sister cried and the old woman said: ``Don't cry, but go up.'' And she climbed up and said: ``My mask is with him.'' And half of his hair was silver and the other half was gold, and his upper teeth were gold, his lower teeth silver.

And the king sent to fetch them, and they descended to fetch them and the old woman cried. And he said: ``Don't cry but come here with us,'' and they brought them up, they went up and he asked: ``You, what kind of human are you, because your shape and appearance resemble a jinn and nymph.''

And he said: ``Great lord, there were two of us, I called the dogs and he, my younger brother, beat sago and I said: `If your brother-in-law comes here, give him sago bread.'

\largerpage
And he came here to ask for it, and she did not give it to him, and he was ashamed and went away, over seven mountains and seven rivers he ventured, and I followed him.

And I found him and after we had finished chewing betel, I said: `Come here, let's return,' and he said: `Well, I'm ashamed to go back; so you return, but I will go on.'{''}

And the king said: ``Well then, take your kris sheath out of your belt, so that I may put my knife in it,'' and he put it into the sheath and it fit. And he the king said: ``Well then, let me put your ring on,'' and he put it on, and it fit. And he said again: ``Well then, let's connect your coconut shells,'' and they connected them and they fit, and the king beat his chest and he said: ``If that's the case, it's me,'' and they celebrated for seven nights and seven days.
