\subsection*{The cuscus}

There was a woman and she gave birth to a cuscus. And the mother asked: ``My cuscus, who shall raise it?'' And of the seven children of the king the youngest wanted to marry it. And the other sisters said: ``Sister, are you so sad over men that you marry a cuscus?'' And she was silent, and then the next day was their feast. And they said: ``Sister, tomorrow we will celebrate during daytime, then your husband shall go down to the shore with his ugly face.''

And she was silent, in this way they celebrated and the cuscus said: ``Strip this coconut of its husk for us, so that we can wash and oil our hair,'' and the older sisters said: ``Sister, so what part of him shall he wash and oil?'' And she dehusked coconuts to wash and oil their hair, they went to the beach and they washed and oiled their hair. And he said: ``You go first,'' and she first washed and oiled her hair. And when she was done with washing and oiling her hair, she said: ``You too,'' and he did it too, and he said: ``Close your eyes,'' and she closed her eyes, and he said: ``It is a mask,'' then he took off his mask.

Half of his hair was silver, the other half was gold, and half of his teeth were silver and the other half was gold. And they returned and they saw him and they, the older sisters, said: ``Sister, our husband'' and he said: ``No, my wife,'' and they said: ``What is he -- our brother-in-law?'' and he said: ``No.''

Then he went away with a troubled heart and to his wife he promised, saying: ``When you give birth, there will be two children, one will be born at the new moon, and one will be born at sunrise.''

Then he left and she gave birth. And the older sisters said: ``We will seal your eyes,'' and they sealed them, and she gave birth. And when her children were born, they brought them outside in a box and let the children float away, after that they put charcoal in her vagina, and coconut bark. And her eyes they rubbed with oil, so they opened again, and the mother asked: ``Sisters, how are the kids?'' And they said: ``Children, where? Here are only charcoal and coconut bark.'' And those children floated landwards to the landing place of the general. And the wife of the general defecated, and she sat down on the box, and in there the children cried.

And she ran inland and she reported, saying: ``Old man, seawards at the landing place inside a piece of wood children are crying,'' and he said: ``Old woman, let's split it open.''

And they split it, and they split it and there were two children.

And now the wife said: ``Old man, let's slaughter them for us,'' and he said: ``Let's not do that, so that we can use them to order them around.'' Then they brought them up until they had grown up.

 Once they smoked one spit human flesh and one spit of pork. And they ate and said, those children: ``Grandfather, there is lots of food that is forbidden to us,'' and he said: ``Old woman, there is a little bit of food left, take it for them, so that our grandchildren may eat the abundant food.''

And they ate, after that he went and called the dogs and he said: ``Old woman, don't let our grandchildren alone, since they are grown up.'' And she said: ``Old man, just go, I won't abandon them.''

And he went, he called the dogs and was gone, and they said: ``Old woman, give us a knife, because seawards by the water we will play.'' And she said: ``Grandchildren, go ahead, play,'' and they played, they went away and they made one proa of stone in the manner of a \textit{juangana}. They made its sails, its masts, its oars, its planks; and then called the old woman, saying: ``Grandchildren, come landwards,'' and they said: ``Old woman, not yet, we are still making our proa.'' And they set up the sail and went away. Then he came back and asked: ``How are our grandchildren?'' And she said: ``Just now they are playing at sea at the landing place.''

And he climbed a \textit{kenari} tree, so he saw their sail as small as a cigarette.

And he extended his pubic hair to bind them but it did not reach the boat, and he said: ``Old woman, pull out your pubic hair so that we can extend mine with it so that we can bind them.'' And he extended it and threw it at them. And they bound them, and following that their sail came closer and closer landwards from the sea.

And they hacked it with an axe but it refused to break, they chopped it but it refused to break and they rubbed it with waste of coconuts, so that, look, mice would gnaw on it, but it refused to break and they came closer and closer to them.

And they went again and they arrived at the landing place of the king and the brother said: ``Put on pants,'' and she put on pants.

And they disembarked and the king said: ``Your sister here, I want to marry her,'' and the brother said: ``Great lord, but we are only men.'' But the king did not believe it, and he, the brother, said: ``Well then, if you don't believe it, let him urinate.'' And he urinated, but he directed the stream through a bamboo joint, so that it reached the ground far away. And he said: ``It's true!'' but afterwards her vagina bled and her pants hurt it. And the king saw it and he said: ``It's not true, it's a woman,'' but a hummingbird came to her head and she caught it with her knife, so blood dripped down on her hand, and she said: ``Here, it is just the blood of the hummingbird,'' and he said: ``It's true.''

And he said: ``It is good, go,'' and they went again, they rowed towards the sea, and she was naked. And he said: ``It is a woman, so you (pl.) [must] row her.'' And they rowed her, they pushed the boat forwards and it sank, ahead, once again, they pushed it and it sank more, ahead, once again, they pushed it and it sank more, ahead, once again, they pushed it and it sank more, and then the boat was gone and only then she said: ``You abandoned it, so we can't use it.''

And they went on, and they came landwards to the landing place of their father, and at sea at his landing place they arrived. And the king said: ``Who is arriving at my landing place?'' and he, the king, said: ``There seawards, fetch them,'' and they went seawards to fetch them and they said: ``We've come from the land seawards,'' the king said: ``Why do they arrive at my landing place?'' And they said: ``If that's the case, let him come seawards,'' and he came seawards and asked: ``Where have you come from?'' And they said: ``We here are the children of King Cuscus. He went with a burdened heart and he promised our mother, saying: `You here shall give birth, and your children will be two, and one will be born at new moon and one at sunrise.' And mother gave birth, and they sealed her eyes and in her vagina they put charcoal and coconut bark and after that the older mothers put us into a box and let us float away. And we arrived landwards at the landing place of the general and he raised us until we had grown up. And he called the dogs and we went, we made our proa of stone.''

And the king said: ``If that's the case, I am your father, so come to shore,'' and they asked: ``How about mother?'' And he said: ``Your mother is by the sea, she is alone, and she will follow us up.''

And they said: ``When we disembark, your six wives must serve as rollers for our proa when we drag it up.''

And the king said: ``That's good,'' and he used them as rollers, so their bones broke.

After that, they fetched the mother and they rubbed her with coconut bark, after that, they washed her beautiful in water, and then they washed her with silver water and and after that with gold water. Then half of her hair was silver and the other half was gold, her teeth were half silver, and the other half was gold.
