\subsection*{Two People who Fasted}

On one day they entered the mosque, they fasted, and the older brother urged the mother, saying: ``Mother, don't slaughter our chicken, even if there is nothing to eat for us, we will return here.''

And the mother said: ``No, we won't strike it dead, just go.'' And when their fasting was finished, they returned.

And the older brother asked: ``Mother, is our chicken still around?'' Then the mother said: ``Hey, children, I slaughtered your chicken,'' and they were angry, they said: ``How is it possible that you slaughtered our chicken?'' And the mother said: ``Because we had nothing to eat, therefore I slaughtered it.'' Then they said: ``Because you slaughtered our chicken, we will go away.''

After that they went and they took provisions for the way: seven watermelons, seven chicken eggs and one whole coconut.

Then they walked on, until they came to a river and there they rested, because night had fallen.

And they lay down to sleep and the older one said: ``In a little while I'll wake you up, then watch carefully, because in a little while, when it is still midnight, a horse will pick us up,'' and the younger one said: ``If that's the case, you must wake me up, don't abandon me.''

And when it was still midnight the horse came, so the older brother woke up the younger one, but he did not wake up.

Then the older brother went, and when the older brother went, he took away the rings of the younger one so that they were exchanged and the older one put his rings on the younger and those of the younger one he put on himself.

The older brother wore seven layers of shirts and pants, and when he came to a crossroads on the way, he dropped his shirt. Then he came to a crossroads again and he dropped one shirt, he came to a crossroads again and he dropped one, until he only had one shirt left and one pair of pants and after that he dropped his kris sheath and it rolled on the left path at the crossroads.

And the older brother followed the rightward path and the younger brother followed the leftward path.

And the younger brother came to an old woman, but first he hid under a banana plant. And one bunch of bananas was ripe, and down there he ate them raw.

Once she went to her garden to harvest her rice, and she went straight to her bananas. But she came there, and of her bananas one bunch was not there anymore; she said: ``Hey, who ate this? Maybe the earwigs.''

And she said: ``If I only find them, then I will strike them dead.'' And he shouted, saying: ``Grandmother!'' Only at the third time she heard him, and she said: ``Hey, who here is calling `grandmother, grandmother?'{''} And he revealed himself, and she came seawards to him and said: ``Hey, my real son.'' Then she cooked rice for him and he ate.
